---
title: An R Markdown document converted from "02_Summaries_and_Groups_ex_solutions.ipynb"
output: html_document
---

# Exercise time:

Do these exercises to `mtcars`-dataset intialized below (do note the use of `rownames_to_column` that will store row names from `data.frame` into a column in `tibble` [[rownames_to_column]](https://tibble.tidyverse.org/reference/rownames.html)):

1. Use `mutate` to calculate litres per 100 km (e.g. lp100km) as mpg is a bad unit to use for fuel consumption (Ref: [[1]](http://science.sciencemag.org/content/320/5883/1593.full?ijkey=3pScQm7pQBzqs&keytype=ref&siteid=sci). Conversion factor is `lp100km=235.214583/mpg`.
2. Calculate correlation of horsepower and quarter mile time (`qsec`). Plot scatter plot and a linear fit.
3. Group data based on cylinder numbers. Calculate correlation between weight and l/100km for each group. Plot scatter plots with different colours for each group.

# Solutions:

```{r}
library(tidyverse)

data(mtcars)

mtcars_tbl <- as_tibble(rownames_to_column(mtcars,var='model'))

str(mtcars_tbl)
```

## 1.

```{r}
mtcars_tbl <- mtcars_tbl %>%
    mutate(lp100km=235.214583/mpg)

str(mtcars_tbl)
```

## 2.

```{r}
mtcars_tbl %>%
    summarize(hp_qsec_cor=cor(hp,qsec))

mtcars_tbl %>%
    ggplot(aes(x=hp,y=qsec)) +
    geom_point() +
    geom_smooth(method=lm)
```

## 3.

```{r}
mtcars_grouped_by_cyl <- mtcars_tbl %>%
    group_by(cyl)

mtcars_grouped_by_cyl %>%
    summarize(wt_lp100km_cor=cor(wt,lp100km))

mtcars_grouped_by_cyl %>%
    ggplot(aes(x=wt,y=lp100km,color=factor(cyl))) +
    geom_point() +
    geom_smooth(method=lm)
```

