---
title: An R Markdown document converted from "06_Multiple_Models_ex_solutions.ipynb"
output: html_document
---

## Exercise

Dataset `/m/jhnas/jupyter/shareddata/python-r-data-analysis/honey-production/honeyproduction.csv` contains data on honey production in the US.

1. Calculate the total amount of honey produced in the US for each year. Plot it.
2. For each state, fit a linear model totalprod ~ year. Plot original data and fitted lines.
3. Plot the slope coefficient of the linear fit.

Loading required packages.

```{r}
library(tidyverse)
if (!file.exists('rlibs')) {
    dir.create('rlibs')
}
if (!file.exists('rlibs/broom')) {
    install.packages('broom', repos="http://cran.r-project.org", lib='rlibs')
}
library(broom, lib.loc='rlibs')
library(modelr)
```

Loading data.

```{r}
honey <- read_csv("/m/jhnas/jupyter/shareddata/python-r-data-analysis/honey-production/honeyproduction.csv")
```

Calculate yearly production using `summarize`.

```{r}
yearly_prod <- honey %>%
    select(year,totalprod) %>%
    group_by(year) %>%
    summarize(totalprod=sum(totalprod))

yearly_prod %>%
    ggplot(aes(x=year, y=totalprod)) +
        geom_col()
```

Create a nested array with the relevant variables.

```{r}
honey_nested <- honey %>%
    select(state, totalprod, year) %>%
    group_by(state) %>%
    nest()
```

Create a model function.

```{r}
honey_model <- function(data) lm(totalprod ~ year, data=data)
```

Fit the models using `map` and `mutate`.

```{r}
honey_nested <- honey_nested %>%
    mutate(
        models=map(data, honey_model)
    )
```

Add predictions using `map2` and `mutate`.

```{r}
honey_nested <- honey_nested %>%
    mutate(
        pred=map2(data, models, add_predictions)
    )
```

Unnest predictions.

```{r}
honey_predictions <- unnest(honey_nested, pred)
head(honey_predictions)
```

Plot predictions.

```{r}
honey_predictions %>%
    ggplot(aes(x=year, y=totalprod, color=state)) +
        geom_point() +
        geom_line(aes(y=pred))
```

Use `map` and `tidy` to obtain model parameters.

```{r}
honey_nested <- honey_nested %>%
    mutate(
        model_params=map(models, tidy)
    )
```

Unnest the model parameters.

```{r}
honey_model_params <- honey_nested %>%
    unnest(model_params)
```

Plot the model parameters.

```{r}
honey_model_params %>%
    filter(term=='year') %>%
    mutate(state=reorder(state,estimate)) %>%
    ggplot(aes(x=state,y=estimate)) +
        geom_col()
```

