---
title: An R Markdown document converted from "01_Introduction_to_R_ex_solutions.ipynb"
output: html_document
---

# Exercise time: 

1. Use `filter` and `geom_histogram` to plot a histogram of ratings for action-movies (Doc pages: [[filter]](http://dplyr.tidyverse.org/reference/filter.html) [[geom_histogram]](http://ggplot2.tidyverse.org/reference/geom_histogram.html)).
2. Use `filter` and `arrange` to find the highest rated romance-movie with more than 1000 votes (Doc pages: [[filter]](http://dplyr.tidyverse.org/reference/filter.html) [[arrange]](https://www.rdocumentation.org/packages/dplyr/versions/0.7.3/topics/arrange)).
3. Take movies with more than 1000 votes and an estimated budget (`! is.na(budget)`). Use `top_n` to limit yourself to 200 highest rated. Use `geom_point` and `scale_x_log10` to create a semi-log scatter plot of movie budget vs. movie rating ([[top_n]](http://dplyr.tidyverse.org/reference/top_n.html) [[geom_point]](http://ggplot2.tidyverse.org/reference/geom_point.html) [[scale_x_log10]](http://ggplot2.tidyverse.org/reference/scale_continuous.html)).

```{r}
library(tidyverse)
library('ggplot2movies', lib.loc='rlibs')
data('movies',package='ggplot2movies')
```

# Solutions:

## 1.

```{r}
movies %>%
    filter(Action == 1) %>%
    ggplot(aes(x=rating)) +
    geom_histogram(binwidth=0.1)
```

```{r}
movies %>%
    filter(Romance==1 & votes > 1000) %>%
    arrange(desc(rating)) %>%
    head(1)
```

```{r}
movies %>%
    filter(votes > 1000 & ! is.na(budget)) %>%
    top_n(rating,n=200) %>%
    ggplot(aes(x=budget,y=rating)) +
    geom_point() +
    scale_x_log10()
```

