---
title: An R Markdown document converted from "06_Multiple_Models.ipynb"
output: html_document
---

# Running a model for split datasets

## Running same model on subsets of a single dataset

This section is based on [Chapter 25](https://r4ds.had.co.nz/many-models.html) of Garrett Grolemund's and Hadley Wickham's excellent book, [R for Data science](https://r4ds.had.co.nz/).

We will be using packages `tidyverse`, `broom`, `modelr` and `gapminder`. `gapminder` is a sample dataset from [Gapminder Foundation](https://www.gapminder.org/).

```{r}
library(tidyverse)
if (!file.exists('rlibs')) {
    dir.create('rlibs')
}
if (!file.exists('rlibs/broom')) {
    install.packages('broom', repos="http://cran.r-project.org", lib='rlibs')
}
if (!file.exists('rlibs/gapminder')) {
    install.packages('gapminder', repos="http://cran.r-project.org", lib='rlibs')
}
library(broom, lib.loc='rlibs')
library(gapminder, lib.loc='rlibs')
library(modelr)
```

`gapminder` dataset contains yearly data of population, life expectancy and GDP on a country by country level.

```{r}
head(gapminder)
```

If we plot life expectancy vs time, we can see that for most countries this relationship seems almost linear.

```{r}
gapminder %>% 
  ggplot(aes(year, lifeExp, group = country)) +
    geom_line(alpha = 1/3)
```

Let's try this approach with one country e.g. New Zealand. Firstly, let's create a new tibble that has only New Zealand's data, Secondly, let's plot it.

```{r}
nz <- filter(gapminder, country == "New Zealand")
nz %>% 
  ggplot(aes(year, lifeExp)) + 
  geom_line() + 
  ggtitle("New Zealand")
```

Let's fit a linear model here.

```{r}
nz_model <- lm(lifeExp ~ year, data = nz)
nz %>% 
  add_predictions(nz_model) %>%
  ggplot(aes(year, pred)) + 
  geom_line() + 
  ggtitle("Linear trend")
```

To see how well this matches our original data, let's add residuals (the difference between fit and data) to the dataset with `add_residuals` [[add_residuals]](https://modelr.tidyverse.org/reference/add_residuals.html).

Let's plot this.

```{r}
nz %>% 
  add_residuals(nz_model) %>% 
  ggplot(aes(year, resid)) + 
  geom_hline(yintercept = 0, colour = "white", size = 3) + 
  geom_line() + 
  ggtitle("Remaining pattern")
```

Not a perfect fit, but not bad either.

When doing the same for all countries, we'll first want to group the data per country and continent.

After that we can use the `nest`-function, to take all data that belongs to said groups into a sub-tibble ([[nest]](http://tidyr.tidyverse.org/reference/nest.html)).

```{r}
by_country <- gapminder %>%
    group_by(country, continent) %>%
    nest()

print(by_country)
```

Now you see that the `data`-variable is of type `list` and for each country we have the data that belongs to said country in its own tibble.

Thus we have split the dataset based on the grouping.

Now we can use `function` to create a generic linear model that we can use for each country.

```{r}
country_model <- function(data) lm(lifeExp ~ year, data = data)
```

Notice that the data that goes into the model is a free parameter that is not yet defined. When running this function, `lm` assumes that the data contains columns `lifeExp` and `year`.

To run this model on each dataset in `data`-column, we can use the `map`-function.

`map` takes a `list` and a `function` and it will apply said `function` for each entry in that `list`. As its output it will return a new `list` with a same number of elements as the original `list`.

By using the `mutate`-function, we can put the resulting `list` into a new column in our tibble.

```{r}
by_country <- by_country %>%
    mutate(
        model=map(data,country_model)
    )
print(by_country)
```

Let's say we want to see how well our models we fitted. For this we'll need to generate the residuals with `add_residuals` [[add_residuals]](https://modelr.tidyverse.org/reference/add_residuals.html).

Add residuals takes two arguments: `data` and a `model`. Thus we'll need to iterate over `data1` and `model1`, `data2` and `model2`, ...

For this we can use the `map2`-function [[map2]](https://purrr.tidyverse.org/reference/map2.html).

As its arguments it takes two lists `x` and `y` and a function `f`. As its output it will return `f(x_i,y_i)` for each element of the lists.

```{r}
by_country <- by_country %>%
    mutate(
        residuals=map2(data, model, add_residuals)
    )

print(by_country)
```

Now we have the residuals in column `residuals`. To obtain the actual values from the `residuals`-column, we'll have to use `unnest` to reverse the nesting procedure [[unnest]](https://tidyr.tidyverse.org/reference/unnest.html).

```{r}
residuals <- unnest(by_country, residuals)
head(residuals)
```

Let's plot the residuals.

```{r}
residuals %>% 
  ggplot(aes(year, resid)) +
    geom_line(aes(group = country), alpha = 1 / 3) + 
    geom_smooth(se = FALSE)
```

Looks ok for the most part. Some countries have quite big residuals.

Let's focus on finding those.

In the previous section we used `tidy` from `broom`-package to obtain parameters for models.

This time, let's use `glance` to obtain statistics on how good the fit was [[glance]](https://broom.tidyverse.org/reference/glance.lm.html).

For example, when used on `nz_model`, we obtain these nice statistics.

```{r}
glance(nz_model)
```

Let's use `map` to do the same on a model-by-model basis.

```{r}
by_country <- by_country %>%
    mutate(
        model_perf=map(model, glance)
    )

print(by_country)
```

Now let's `unnest` the data.

```{r}
model_perf <- unnest(by_country, model_perf)
print(model_perf)
```

This left the columns with `list`-datatype in the `model_perf`-tibble. Let's forcibly drop them with `.drop=TRUE` during unnesting.

```{r}
model_perf <- unnest(by_country, model_perf, .drop=TRUE)
head(model_perf)
```

Now we have the fit statistics in a nice tidy format. Let's find the countries that bucked the trends by looking at the worst `r.squared` scores. 

```{r}
bad_fit <- filter(model_perf, r.squared < 0.25)

gapminder %>% 
  semi_join(bad_fit, by = "country") %>% 
  ggplot(aes(year, lifeExp, colour = country)) +
    geom_line()
```

## Exercise

Dataset `/m/jhnas/jupyter/shareddata/python-r-data-analysis/honey-production/honeyproduction.csv` contains data on honey production in the US.

1. Calculate the total amount of honey produced in the US for each year. Plot it.
2. For each state, fit a linear model totalprod ~ year. Plot original data and fitted lines.
3. Plot the slope coefficient of the linear fit.

```{r}
honey <- read_csv("/m/jhnas/jupyter/shareddata/python-r-data-analysis/honey-production/honeyproduction.csv")
```

