---
title: An R Markdown document converted from "07_Bringing_It_Home_ex_solutions.ipynb"
output: html_document
---

# Exercise

It is often said that increase in life expectancy will reduce the amount of children people have as the probability of the child living into adulthood will be higher.

There's two files in `/m/jhnas/jupyter/shareddata/python-r-data-analysis/world-bank-data-1960-to-2016`:
    
    - `fertility_rate.csv` contains fertility rate data.
    - `life_expectancy.csv` constains life expectancy data.

There's one file in `/m/jhnas/jupyter/shareddata/python-r-data-analysis/world-capitals-gps`:

    - `concap.csv` contains GPS locations of world's capitals (and their country codes and continent codes).

Do the following:

1. Read these datasets in and make them into tidy format.

2. Use join-functions from `tidyverse` to join the important variables of these datasets together [[join]](https://dplyr.tidyverse.org/reference/join.html).

3. Calculate mean values and variations of fertility and life expectancy on a global and continental scale.

4. Fit linear model of life expectancy vs. fertility to each country. See which countries match the best and which do the worst.

5. Fit linear model of year vs life expectancy to each country. See which countries have done the most to improve on life expectancy since 1960.

```{r}
library(tidyverse)
if (!file.exists('rlibs')) {
    dir.create('rlibs')
}
if (!file.exists('rlibs/broom')) {
    install.packages('broom', repos="http://cran.r-project.org", lib='rlibs')
}
library(broom, lib.loc='rlibs')
library(modelr)
```

```{r}
fertility <- read_csv('/m/jhnas/jupyter/shareddata/python-r-data-analysis/world-bank-data-1960-to-2016/fertility_rate.csv') %>%
    gather('1960':'2016', key='year', value='fertility')%>%
    select(-"Indicator Name",-"Indicator Code", -"Country Code") %>%
    rename("country"="Country Name")
head(fertility)
```

```{r}
life_expectancy <- read_csv('/m/jhnas/jupyter/shareddata/python-r-data-analysis/world-bank-data-1960-to-2016/life_expectancy.csv') %>%
    gather('1960':'2016', key='year', value='life_expt')%>%
    select(-"Indicator Name",-"Indicator Code", -"Country Code") %>%
    rename("country"="Country Name")
head(life_expectancy)
```

```{r}
continents <- read_csv('/m/jhnas/jupyter/shareddata/python-r-data-analysis/world-capitals-gps/concap.csv') %>%
    rename("country"=CountryName) %>%
    rename("continent"=ContinentName) %>%
    select(country, continent)
continents
```

```{r}
all_data <- fertility %>%
    left_join(life_expectancy) %>%
    left_join(continents) %>%
    drop_na() %>%
    mutate(year=as.integer(year))

head(all_data)
```

```{r}
all_data %>%
    summarize_at(vars(life_expt, fertility), list(mean=mean, var=var)) %>%
    arrange(desc(life_expt_mean))
```

```{r}
all_data %>%
    group_by(continent) %>%
    summarize_at(vars(life_expt, fertility), list(mean=mean, var=var)) %>%
    arrange(desc(life_expt_mean))
```

```{r}
fertility_model <- function(data) lm(fertility ~ life_expt, data=data)
```

```{r}
countries <- all_data %>%
    group_by(continent, country) %>%
    nest()
print(countries)
```

```{r}
countries <- countries %>%
    mutate(
        fert_models=map(data, fertility_model)
    )
print(countries)
```

```{r}
countries <- countries %>%
    mutate(
        fert_residuals=map2(data, fert_models, add_residuals)
    )
```

```{r}
residuals <- unnest(countries, fert_residuals)

residuals %>%
    ggplot(aes(x=year, y=resid)) +
    geom_line(aes(group=country), alpha=1/3) +
    geom_smooth(se = FALSE)
```

```{r}
countries <- countries %>%
    mutate(
        fert_stats=map(fert_models, glance)
    )
```

```{r}
fert_stats <- unnest(countries, fert_stats, .drop=TRUE)

fert_top5 <- fert_stats %>%
    top_n(5, w=r.squared)

fert_bot5 <- fert_stats %>%
    top_n(-5, w=r.squared)

fert_top5
fert_bot5
```

```{r}
countries <- countries %>%
    mutate(
        fert_pred=map2(data, fert_models, add_predictions)
    )
```

```{r}
fert_preds <- unnest(countries, fert_pred, .drop=TRUE)

fert_top5 <- fert_top5 %>%
    left_join(fert_preds) %>%
    drop_na()

fert_bot5 <- fert_bot5 %>%
    left_join(fert_preds) %>%
    drop_na()
```

```{r}
fert_top5 %>%
    ggplot(aes(x=life_expt, y=fertility, color=country)) +
        geom_point() +
        geom_line(aes(y=pred))

fert_bot5 %>%
    ggplot(aes(x=life_expt, y=fertility, color=country)) +
        geom_point() +
        geom_line(aes(y=pred))
```

```{r}
lxpt_model <- function(data) lm(life_expt ~ year, data=data)
```

```{r}
countries <- countries %>%
    mutate(
        lxpt_models=map(data, lxpt_model),
        lxpt_params=map(lxpt_models, tidy),
        lxpt_pred=map2(data, lxpt_models, add_predictions)
    )
print(countries)
```

```{r}
lxpt_params <- unnest(countries, lxpt_params, .drop=TRUE)
head(lxpt_params)

lxpt_preds <- unnest(countries, lxpt_pred, .drop=TRUE)
```

```{r}
lxpt_top5 <- lxpt_params %>%
    filter(term=='year') %>%
    top_n(5,w=estimate)

lxpt_bot5 <- lxpt_params %>%
    filter(term=='year') %>%
    top_n(-5,w=estimate)
```

```{r}
lxpt_top5 <- lxpt_top5 %>%
    left_join(lxpt_preds) %>%
    drop_na()

lxpt_bot5 <- lxpt_bot5 %>%
    left_join(lxpt_preds) %>%
    drop_na()
```

```{r}
lxpt_top5 %>%
    ggplot(aes(x=year, y=life_expt, color=country)) +
        geom_point() +
        geom_line(aes(y=pred))

lxpt_bot5 %>%
    ggplot(aes(x=year, y=life_expt, color=country)) +
        geom_point() +
        geom_line(aes(y=pred))
```

